A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-22 paper.Rmd
up to 2.3

We can assess residuals in multiple ways. To examine the their univariate distribution, residuals may be plotted as a histogram or normal probability plot. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, we would consider it to be well-behaved. However, if the distribution is skewed, bimodal, multimodal, or contains outliers, there would be cause for concern. We can also inspect the distribution by conducting a goodness-of-fit test, such as the Shapiro-Wilk (SW) normality test [@shapiro1965analysis].

@@

Graphical summaries where residuals are plotted against fitted values, or other functions of the predictors (expected to be approximately orthogonal to the residuals) are considered to be the most important residual plots by @cook_applied_1999. Figure \ref{fig:residual-plot-common-departures}A shows an example of an ideal residual plot where points are symmetrically distributed around the horizontal zero line (red), with no discernible patterns. There can be various types of departures from this ideal pattern. Non-linearity, heteroskedasticity and non-normality are three commonly checked departures.

@@

Model misspecification occurs if functions of predictors that needed to accurately describe the relationship between response and predictors were omitted. Any non-linear pattern visible in the residual plot could be indicative of this problem. An example residual plot containing visual pattern of non-linearity is shown in Figure \ref{fig:residual-plot-common-departures}B. One can clearly observe the "S-shape" from the residual plot, which corresponds to cubic term that should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D) shows a scatterplot where the residuals have a skewed distribution, as seen by the uneven vertical spread. Unlike non-linearity and heteroskedasticity, non-normality is usually detected with a different type of residual plot, a histogram or normal probability plot.  Because we focus on scatterplots, non-normality is not one of the departures examined in depth in this paper.

@@

Many different hypothesis tests are available to detect specific model defects. For example, the presence of heteroskedasticity can usually be tested by applying the White test [@white_heteroskedasticity-consistent_1980] or the BP test [@breusch_simple_1979], which are both derived from the Lagrange multiplier test [@silvey1959lagrangian] principle that relies on the asymptotic properties of the null distribution. To test specific forms of non-linearity, one may apply the F-test as a model structural test to examine the significance of specific polynomial and non-linear forms of the predictors, or the significance of proxy variables as in the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey_tests_1969]. The SW test [@shapiro1965analysis] is the most widely used test of non-normality included by many of the statistical software programs. The Jarque--Bera test [@jarque1980efficient] is also used to directly check whether the sample skewness and kurtosis match a normal distribution.

Table \ref{tab:example-residual-plot-table} displays the $p$-values from the RESET, BP and SW tests applied to the residual plots in Figure \ref{fig:residual-plot-common-departures}. The RESET test and BP test were computed using the `resettest` and `bptest` functions from the R package `lmtest`, respectively. The SW test was computed using the `shapiro.test` from the core R package `stats`. (The the R package `skedastic` [@skedastic] contains a large collection of tests for heteroskedasticity.) Although, the RESET test is exact, it requires the selection of a power parameter. According to @ramsey_tests_1969, a power of four is recommended, which we adopted in our analysis.  The BP and SW tests are approximate.

We would expect the RESET test for non-linearity to reject residual plot B, the BP test for heteroskedasticity to reject the residual plot C, and SW test for non-normality to reject residual plot D, which they all do and correctly fail to reject residual plot A. Interestingly, the BP and SW tests also reject the residual plots exhibiting structure that they weren't designed for. @cook1982residuals discusses that most residual-based tests for particular types of departure from model assumptions are also sensitive to other types of departures. This could be considered a Type III error, where the null hypothesis of good residuals is correctly rejected but for the wrong reason. Also, some types of departure can have elements of other types of departure, for example, non-linearity could be viewed as heteroskedasticity. Additionally, other problesms such as outliers can trigger the rejection [@cook_applied_1999].

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (This is common knowledge but a good discussion can be found in @kirk1996.) While such rejections may be statistically correct, their sensitivity may render the results impractical. This is especially pertinent for residual diagnostics. Therefore, a key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses. However, minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. In fact, the experiment discussed in this paper specifically addresses this.

>>>>

We can assess residuals in multiple ways. To examine the their univariate distribution, residuals may be plotted as a histogram or normal probability plot. Using the classical normal linear regression model as an example, if the distribution is symmetric and unimodal, we would consider it to be well-behaved. However, if the distribution is skewed, bimodal, multimodal, or contains outliers, there would be cause for concern. We can also inspect the distribution by conducting a goodness-of-fit test, such as the Shapiro-Wilk normality test [@shapiro1965analysis].

@@

Graphical summaries where residuals are plotted against fitted values, or other functions of the predictors (expected to be approximately orthogonal to the residuals) are considered to be the most important residual plots by @cook_applied_1999. Figure \ref{fig:residual-plot-common-departures}A shows an example of an ideal residual plot where points are symmetrically distributed around the horizontal zero line (red), with no discernible patterns. There can be various types of departures from this ideal pattern. Non-linearity, heteroskedasticity and non-normality, shown in Figure \ref{fig:residual-plot-common-departures}B,C,D are three commonly checked departures.

@@

Model misspecification occurs if functions of predictors that needed to accurately describe the relationship between response and predictors are omitted. Any non-linear pattern visible in the residual plot could be indicative of this problem. An example residual plot containing visual pattern of non-linearity is shown in Figure \ref{fig:residual-plot-common-departures}B. One can clearly observe the "S-shape" from the residual plot, which corresponds to cubic term that should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D) shows a scatterplot where the residuals have a skewed distribution, as seen by the uneven vertical spread. Unlike non-linearity and heteroskedasticity, non-normality is usually detected with a different type of residual plot: a histogram or normal probability plot.  Because we focus on scatterplots, non-normality is not one of the departures examined in depth in this paper. (@loy_variations_2016 and @loy2013diagnostic discuss related work on non-normality checking.)

@@

Many different hypothesis tests are available to detect specific model defects. For example, the presence of heteroskedasticity can usually be tested by applying the White test [@white_heteroskedasticity-consistent_1980] or the Breusch-Pagan (BP) test [@breusch_simple_1979], which are both derived from the Lagrange multiplier test [@silvey1959lagrangian] principle that relies on the asymptotic properties of the null distribution. To test specific forms of non-linearity, one may apply the F-test as a model structural test to examine the significance of specific polynomial and non-linear forms of the predictors, or the significance of proxy variables as in the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey_tests_1969]. The Shapiro-Wilk (SW) normality test [@shapiro1965analysis] is the most widely used test of non-normality included by many of the statistical software programs. The Jarque-Bera test [@jarque1980efficient] is also used to directly check whether the sample skewness and kurtosis match a normal distribution.

Table \ref{tab:example-residual-plot-table} displays the $p$-values from the RESET, BP and SW tests applied to the residual plots in Figure \ref{fig:residual-plot-common-departures}. The RESET test and BP test were computed using the `resettest` and `bptest` functions from the R package `lmtest`, respectively. The SW test was computed using the `shapiro.test` from the core R package `stats`. (Although we didn't use it, it's useful to know that the R package `skedastic` [@skedastic] also contains a large collection of functions to test for heteroskedasticity.) Although, the RESET test is exact, it requires the selection of a power parameter.  @ramsey_tests_1969 recommends a power of four, which we adopted in our analysis.  The BP and SW tests are approximate.

We would expect the RESET test for non-linearity to reject residual plot B, the BP test for heteroskedasticity to reject the residual plot C, and the SW test for non-normality to reject residual plot D, which they all do and all tests also correctly fail to reject residual plot A. Interestingly, the BP and SW tests also reject the residual plots exhibiting structure that they weren't designed for. @cook1982residuals explain that most residual-based tests for particular types of departure from model assumptions are also sensitive to other types of departures. This could be considered a Type III error (XXXREF), where the null hypothesis of good residuals is correctly rejected but for the wrong reason. Also, some types of departure can have elements of other types of departure, for example, non-linearity could be viewed as heteroskedasticity. Additionally, other data problems such as outliers can trigger rejection (or not) of the null hypothesis [@cook_applied_1999].

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (A good discussion can be found in @kirk1996.) While such rejections may be statistically correct, their sensitivity may render the results impractical. A key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses, but minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. The experiment discussed in this paper specifically addresses this tension between statistical and practical statistics.

### 2023-04-22 paper.tex
up to 2.3

We can assess residuals in multiple ways. To examine the their
univariate distribution, residuals may be plotted as a histogram or
normal probability plot. Using the classical normal linear regression
model as an example, if the distribution is symmetric and unimodal, we
would consider it to be well-behaved. However, if the distribution is
skewed, bimodal, multimodal, or contains outliers, there would be cause
for concern. We can also inspect the distribution by conducting a
goodness-of-fit test, such as the Shapiro-Wilk (SW) normality test
\citep{shapiro1965analysis}.

@@

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are considered to be the most important
residual plots by \citet{cook_applied_1999}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality are three commonly checked
departures.

@@

Model misspecification occurs if functions of predictors that needed to
accurately describe the relationship between response and predictors
were omitted. Any non-linear pattern visible in the residual plot could
be indicative of this problem. An example residual plot containing
visual pattern of non-linearity is shown in Figure
\ref{fig:residual-plot-common-departures}B. One can clearly observe the
``S-shape'' from the residual plot, which corresponds to cubic term that
should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D) shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot, a histogram or normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper.

@@

Many different hypothesis tests are available to detect specific model
defects. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the BP test
\citep{breusch_simple_1979}, which are both derived from the Lagrange
multiplier test \citep{silvey1959lagrangian} principle that relies on
the asymptotic properties of the null distribution. To test specific
forms of non-linearity, one may apply the F-test as a model structural
test to examine the significance of specific polynomial and non-linear
forms of the predictors, or the significance of proxy variables as in
the Ramsey Regression Equation Specification Error Test (RESET)
\citep{ramsey_tests_1969}. The SW test \citep{shapiro1965analysis} is
the most widely used test of non-normality included by many of the
statistical software programs. The Jarque--Bera test
\citep{jarque1980efficient} is also used to directly check whether the
sample skewness and kurtosis match a normal distribution.

Table \ref{tab:example-residual-plot-table} displays the \(p\)-values
from the RESET, BP and SW tests applied to the residual plots in Figure
\ref{fig:residual-plot-common-departures}. The RESET test and BP test
were computed using the \texttt{resettest} and \texttt{bptest} functions
from the R package \texttt{lmtest}, respectively. The SW test was
computed using the \texttt{shapiro.test} from the core R package
\texttt{stats}. (The the R package \texttt{skedastic} \citep{skedastic}
contains a large collection of tests for heteroskedasticity.) Although,
the RESET test is exact, it requires the selection of a power parameter.
According to \citet{ramsey_tests_1969}, a power of four is recommended,
which we adopted in our analysis. The BP and SW tests are approximate.

We would expect the RESET test for non-linearity to reject residual plot
B, the BP test for heteroskedasticity to reject the residual plot C, and
SW test for non-normality to reject residual plot D, which they all do
and correctly fail to reject residual plot A. Interestingly, the BP and
SW tests also reject the residual plots exhibiting structure that they
weren't designed for. \citet{cook1982residuals} discusses that most
residual-based tests for particular types of departure from model
assumptions are also sensitive to other types of departures. This could
be considered a Type III error, where the null hypothesis of good
residuals is correctly rejected but for the wrong reason. Also, some
types of departure can have elements of other types of departure, for
example, non-linearity could be viewed as heteroskedasticity.
Additionally, other problesms such as outliers can trigger the rejection
\citep{cook_applied_1999}.

With large sample sizes, hypothesis tests may reject the null hypothesis
when there is only a small effect. (This is common knowledge but a good
discussion can be found in \citet{kirk1996}.) While such rejections may
be statistically correct, their sensitivity may render the results
impractical. This is especially pertinent for residual diagnostics.
Therefore, a key goal of residual plot diagnostics is to identify
potential issues that could lead to incorrect conclusions or errors in
subsequent analyses. However, minor defects in the model are unlikely to
have a significant impact and may be best disregarded for practical
purposes. In fact, the experiment discussed in this paper specifically
addresses this.

>>>>

We can assess residuals in multiple ways. To examine the their
univariate distribution, residuals may be plotted as a histogram or
normal probability plot. Using the classical normal linear regression
model as an example, if the distribution is symmetric and unimodal, we
would consider it to be well-behaved. However, if the distribution is
skewed, bimodal, multimodal, or contains outliers, there would be cause
for concern. We can also inspect the distribution by conducting a
goodness-of-fit test, such as the Shapiro-Wilk normality test
\citep{shapiro1965analysis}.

@@

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are considered to be the most important
residual plots by \citet{cook_applied_1999}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality, shown in Figure
\ref{fig:residual-plot-common-departures}B,C,D are three commonly
checked departures.

@@

Model misspecification occurs if functions of predictors that needed to
accurately describe the relationship between response and predictors are
omitted. Any non-linear pattern visible in the residual plot could be
indicative of this problem. An example residual plot containing visual
pattern of non-linearity is shown in Figure
\ref{fig:residual-plot-common-departures}B. One can clearly observe the
``S-shape'' from the residual plot, which corresponds to cubic term that
should have been included in the model.

@@

Figure \ref{fig:residual-plot-common-departures}D) shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot: a histogram or normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper. (\citet{loy_variations_2016} and
\citet{loy2013diagnostic} discuss related work on non-normality
checking.)

@@

Many different hypothesis tests are available to detect specific model
defects. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the Breusch-Pagan
(BP) test \citep{breusch_simple_1979}, which are both derived from the
Lagrange multiplier test \citep{silvey1959lagrangian} principle that
relies on the asymptotic properties of the null distribution. To test
specific forms of non-linearity, one may apply the F-test as a model
structural test to examine the significance of specific polynomial and
non-linear forms of the predictors, or the significance of proxy
variables as in the Ramsey Regression Equation Specification Error Test
(RESET) \citep{ramsey_tests_1969}. The Shapiro-Wilk (SW) normality test
\citep{shapiro1965analysis} is the most widely used test of
non-normality included by many of the statistical software programs. The
Jarque-Bera test \citep{jarque1980efficient} is also used to directly
check whether the sample skewness and kurtosis match a normal
distribution.

Table \ref{tab:example-residual-plot-table} displays the \(p\)-values
from the RESET, BP and SW tests applied to the residual plots in Figure
\ref{fig:residual-plot-common-departures}. The RESET test and BP test
were computed using the \texttt{resettest} and \texttt{bptest} functions
from the R package \texttt{lmtest}, respectively. The SW test was
computed using the \texttt{shapiro.test} from the core R package
\texttt{stats}. (Although we didn't use it, it's useful to know that the
R package \texttt{skedastic} \citep{skedastic} also contains a large
collection of functions to test for heteroskedasticity.) Although, the
RESET test is exact, it requires the selection of a power parameter.
\citet{ramsey_tests_1969} recommends a power of four, which we adopted
in our analysis. The BP and SW tests are approximate.

We would expect the RESET test for non-linearity to reject residual plot
B, the BP test for heteroskedasticity to reject the residual plot C, and
the SW test for non-normality to reject residual plot D, which they all
do and all tests also correctly fail to reject residual plot A.
Interestingly, the BP and SW tests also reject the residual plots
exhibiting structure that they weren't designed for.
\citet{cook1982residuals} explain that most residual-based tests for
particular types of departure from model assumptions are also sensitive
to other types of departures. This could be considered a Type III error
(XXXREF), where the null hypothesis of good residuals is correctly
rejected but for the wrong reason. Also, some types of departure can
have elements of other types of departure, for example, non-linearity
could be viewed as heteroskedasticity. Additionally, other data problems
such as outliers can trigger rejection (or not) of the null hypothesis
\citep{cook_applied_1999}.

With large sample sizes, hypothesis tests may reject the null hypothesis
when there is only a small effect. (A good discussion can be found in
\citet{kirk1996}.) While such rejections may be statistically correct,
their sensitivity may render the results impractical. A key goal of
residual plot diagnostics is to identify potential issues that could
lead to incorrect conclusions or errors in subsequent analyses, but
minor defects in the model are unlikely to have a significant impact and
may be best disregarded for practical purposes. The experiment discussed
in this paper specifically addresses this tension between statistical
and practical statistics.
